---
title: "STANCE_breast_cancer"
author: "Nikki Xiao"
date: "2025-10-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## STANCE

load in data

```{r}
library(SeuratObject)
library(dplyr)
library(Seurat)
#library(Matrix)
# load in data
load(paste0("/Users/nikkixiaomengqi/Desktop/25_Summer_Research/breast_cancer_visium_img_updated.rdata"))
raw_counts <- GetAssayData(visium_data, assay = "RNA", slot = "counts")

# Set RNA as default assay
DefaultAssay(visium_data) <- "RNA"

# Basic information
print(visium_data)
print(paste("Number of spots:", ncol(visium_data)))
print(paste("Total features:", nrow(visium_data)))

# View available assays
print("Available assays:")
print(Assays(visium_data))

# Summary
print("=== Visium Data Summary ===")
print(paste("Total spots:", ncol(visium_data)))
print(paste("RNA features:", nrow(visium_data[["RNA"]])))
```

```{r}
# installed STANCE
library(STANCE)
#library(Matrix)
library(ggplot2)
library(reshape2)
library(dplyr)
```

### **Load in data**

```{r}
library(Seurat)
# load in data
load(paste0("/Users/nikkixiaomengqi/Desktop/25_Summer_Research/breast_cancer_visium_img_updated.rdata"))

raw_counts <- GetAssayData(visium_data, assay = "RNA", slot = "counts")
DefaultAssay(visium_data) <- "RNA"

print("=== Visium Data Summary ===")
print(paste("Total spots:", ncol(visium_data)))
print(paste("RNA features:", nrow(visium_data[["RNA"]])))

visium_data$nCount_RNA[1:5]
```

```{r}
counts_visium <- raw_counts # RNA: genes x spots

img <- Images(visium_data)[1]
coords_df <- GetTissueCoordinates(visium_data, image = img)
rownames(coords_df) <- coords_df$cell
pos_mat <- as.matrix(coords_df[, c("x", "y")])
colnames(pos_mat) <- c("x", "y")

```

```{r}
# load CARD result
load("/Users/nikkixiaomengqi/Desktop/25_Summer_Research/CARD_result_w_Wu2021.rdata")

new_coord <- CARD_obj@spatial_location
colnames(new_coord) <- c("y", "x")

proportion_matrix <- CARD_obj@Proportion_CARD

# Combine the two columns
tumor_epithelial_cells <- proportion_matrix[, "Normal Epithelial"] + proportion_matrix[, "Cancer Epithelial"]

# Remove the original columns
proportion_matrix <- proportion_matrix[, !colnames(proportion_matrix) %in% c("Normal Epithelial","Cancer Epithelial")]

# Add the combined column
proportion_matrix <- cbind(proportion_matrix, "Tumor or Normal Epithelial" = tumor_epithelial_cells)

# subset to the 3 targets
targets <- c("T-cells","B-cells","Tumor or Normal Epithelial")

prop3 <- proportion_matrix[, targets, drop = FALSE]

# Align rows to Visium object
common_spots <- intersect(rownames(prop3), colnames(visium_data))
prop3 <- prop3[common_spots, , drop = FALSE]
visium_data <- visium_data[, common_spots]

```

```{r}
prop3_tcells <- prop3[ , "T-cells", drop = FALSE]
head(prop3_tcells)

prop3_tcells <- as.matrix(prop3[, "T-cells", drop = FALSE])
colnames(prop3_tcells) <- "T-cells"

common <- Reduce(intersect, list(colnames(counts_visium),
                                 rownames(prop3_tcells),
                                 rownames(pos_mat)))

counts_visium2  <- counts_visium[, common, drop = FALSE]
pos_mat2        <- pos_mat[common, , drop = FALSE]
prop3_tcells2   <- prop3_tcells[common, , drop = FALSE]
```

```{r}
# create stance obj
obj_st <- creatSTANCEobject(counts = counts_visium,
                         pos = pos_mat,
                         prop = prop3,
                         covariates = NULL)
save(obj_st, file = "stance_object.rdata")
```

```{r}

# 1) Coerce counts to a plain numeric matrix
counts_mat <- as.matrix(counts_visium)
storage.mode(counts_mat) <- "double"  # ensure numeric

# (quick sanity)
is.matrix(counts_mat); is.numeric(counts_mat)
# --- Quick STANCE preflight checks ---
list(
  # types & shapes
  prop_is_matrix      = is.matrix(prop_t),
  prop_nrow_ncol      = dim(prop_t),
  prop_colnames       = colnames(prop_t),
  prop_range          = range(prop_t, na.rm = TRUE),
  prop_has_na         = anyNA(prop_t),

  counts_is_matrix    = is.matrix(counts_mat),
  counts_mode_numeric = is.numeric(counts_mat),
  counts_dim          = dim(counts_mat),

  pos_is_matrix       = is.matrix(pos_mat),
  pos_dim             = dim(pos_mat),
  pos_colnames        = colnames(pos_mat),

  # alignment (these should be TRUE)
  align_prop_counts   = identical(rownames(prop_t),  colnames(counts_mat)),
  align_pos_counts    = identical(rownames(pos_mat), colnames(counts_mat)),

  # intersection size (should equal ncol(counts) if aligned)
  common_len          = length(Reduce(intersect, list(
                           colnames(counts_mat),
                           rownames(pos_mat),
                           rownames(prop_t))))
)

```

```{r}

#common <- Reduce(intersect, list(colnames(counts_visium), rownames(pos_mat), rownames(prop_t)))

obj_st <- creatSTANCEobject(
  counts = counts_mat,   # <- use the coerced numeric matrix
  pos    = pos_mat,
  prop   = prop_t,       # 1-column matrix, colname "T-cells"
  covariates = NULL
)
```

```{r}
## 0) Counts as plain numeric matrix
counts_mat <- as.matrix(counts_visium)
storage.mode(counts_mat) <- "double"

## 1) Build a 2-col composition: T-cells and "Other" = 1 - T-cells
tvec <- prop3[, "T-cells"]
other <- pmax(0, 1 - tvec)  # keep in [0,1]; OK even if original sums ≈1

prop_two <- cbind("T-cells" = tvec, "Other" = other)
rownames(prop_two) <- rownames(prop3)  # spots

## 2) Align spots (safe)
common <- Reduce(intersect, list(colnames(counts_mat), rownames(pos_mat), rownames(prop_two)))
counts_m <- counts_mat[, common, drop = FALSE]
pos_m    <- as.matrix(pos_mat[common, , drop = FALSE])
prop_m   <- prop_two[common, , drop = FALSE]

## 3) Create STANCE object (now prop is guaranteed a matrix with >=2 cols)
obj_st <- creatSTANCEobject(
  counts = counts_m,
  pos    = pos_m,
  prop   = prop_m,
  covariates = NULL
)

```

Preprocess the input data

```{r}
#load("/Users/nikkixiaomengqi/Desktop/25_Summer_Research/stance_object.rdata")
mySTANCE <- data_preprocess(object = obj_st, 
                            gene.threshold = 0.05, spot.threshold = 10,
                            normalized = FALSE)
```

build gaussian kernel

```{r}
mySTANCE <- build_kernelMatrix(object = mySTANCE)
```

## **Perform STANCE overall test for utSVGs (including both SVGs & ctSVGs)**

```{r}
mySTANCE <- runTest1(object = mySTANCE,
                     correction = F, pv.adjust = "BY")
```

```{r}
head(mySTANCE@Test_1)
```

## **Perform STANCE individual test**

```{r}
genes.list <- rownames(mySTANCE@gene_expression)
# List for utSVGs detected by STANCE overall test
utSVG.list <- genes.list[mySTANCE@Test_1$p_value_adj < 0.05]
utSVG.list[171:180]
```

```{r}

mySTANCE <- runTest2(object = mySTANCE, 
                     Genes_to_test = utSVG.list[1:200], 
                     Cell_types_to_test = "T-cells", 
                     correction = F, 
                     ncores = 1)
```

```{r}
save(mySTANCE, file = "1-200_STANCE_Test2")
```

```{r}
mySTANCE <- runTest2(object = mySTANCE, 
                     Genes_to_test = utSVG.list[201:300], 
                     Cell_types_to_test = "T-cells", 
                     correction = F, 
                     ncores = 1)
```

```{r}
save(mySTANCE, file = "201-300_STANCE_Test2")
```

```{r}
mySTANCE <- runTest2(object = mySTANCE, 
                     Genes_to_test = utSVG.list[301:400], 
                     Cell_types_to_test = "T-cells", 
                     correction = F, 
                     ncores = 1)
```

```{r}
save(mySTANCE, file = "301-400_STANCE_Test2")
```

```{r}
mySTANCE <- runTest2(object = mySTANCE, 
                     Genes_to_test = utSVG.list[401:500], 
                     Cell_types_to_test = "T-cells", 
                     correction = F, 
                     ncores = 1)
```

```{r}
save(mySTANCE, file = "401-500_STANCE_Test2")
```

```{r}
# Individual testing results for cell type "T-cells"
head(mySTANCE@Test_2[["T-cells"]])
```

## Merge results, should be ready to visualize and select top genes

load in Test2 results

```{r}
load("1-200_STANCE_Test2")
p1 <- mySTANCE
load("201-300_STANCE_Test2")
p2 <- mySTANCE
load("301-400_STANCE_Test2")
p3 <- mySTANCE
load("401-500_STANCE_Test2")
p4 <- mySTANCE

stance_obj <- list(p1, p2, p3, p4)

```

merge results

```{r}
celltypes <- names(stance_obj[[1]]@Test_2) # "T-cells"

merged_Test2 <- lapply(celltypes, function(ct) {
  # Extract and combine for this cell type
  dfs <- lapply(stance_obj, function(obj) obj@Test_2[[ct]])
  df_all <- do.call(rbind, dfs)
  df_all <- df_all[!duplicated(rownames(df_all)), , drop = FALSE]
  df_all <- df_all[order(df_all$p_value), , drop = FALSE]
  return(df_all)
})
names(merged_Test2) <- celltypes

# Attach back to one object (e.g., the first)
merged_obj <- stance_obj[[1]]
merged_obj@Test_2 <- merged_Test2
```

Add adjusted p-value

```{r}
# Apply FDR correction (Benjamini–Hochberg) to each cell type
merged_Test2 <- lapply(merged_Test2, function(df) {
  # Compute adjusted p-values (BH FDR)
  df$p_value_adj <- p.adjust(df$p_value, method = "BH")
  
  # Sort by adjusted p-value
  df <- df[order(df$p_value_adj), , drop = FALSE]
  return(df)
})

# Re-attach to your merged STANCE object
merged_obj@Test_2 <- merged_Test2



```

```{r}
# loop through each cell type and append an adjusted p-value column
for (ct in names(merged_obj@Test_2)) {
  df <- merged_obj@Test_2[[ct]]                # extract existing table
  df$p_value_adj <- p.adjust(df$p_value, method = "BH")  # add FDR-adjusted p-value
  merged_obj@Test_2[[ct]] <- df                # put it back
}
```

```{r}
head(merged_obj@Test_2$`T-cells`)
```

```{r}
top_T <- head(rownames(merged_obj@Test_2$`T-cells`), 10)
top_T
```

Volcano Plot

```{r}
# Loading relevant libraries 
library(conflicted)
library(dplyr)
library(tidyverse) # includes ggplot2, for data visualisation. dplyr, for data manipulation.
library(RColorBrewer) # for a colourful plot
library(ggrepel) # for nice annotations
```

```{r}
for (ct in names(merged_obj@Test_2)) {
  df <- merged_obj@Test_2[[ct]]
  
  # add -log10 of adjusted p-value
  df$neglog10_padj <- -log10(df$p_value_adj)

  # put it back
  merged_obj@Test_2[[ct]] <- df
}
head(df)
```

```{r}

```

```{r}
sig_color <- "firebrick3"
up_color <- "steelblue3"
down_color <- "darkgrey"
cutoff <- 0.5
alpha <- 0.05

df <- df %>%
  mutate(
    sig = case_when(
      p_value < alpha & statistic > cutoff  ~ "Up",
      p_value < alpha & statistic < cutoff ~ "Down",
      TRUE                                     ~ "NS"  # not significant
    )
  )

top_label <- df %>%
  arrange(p_value_adj) %>%
  slice_head(n = 10) %>%
  mutate(Gene = rownames(.)) 

p1 <- ggplot(df, aes(x = statistic, y = -log10(p_value))) +
  geom_point(aes(color = sig), alpha = 0.8, size = 1.5) +
  scale_color_manual(values = c("Up" = "firebrick3",
                                "Down" = "steelblue3",
                                "NS" = "grey70")) +
  geom_vline(xintercept = c(cutoff, cutoff),
             linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = -log10(alpha),
             linetype = "dashed", color = "grey40") +
  geom_text_repel(
    data = top_label,
    aes(label = Gene),
    size = 3,
    color = "black",
    box.padding = 0.4,
    segment.color = "grey60",
    max.overlaps = Inf
  ) +
  labs(
    title = paste0("STANCE Volcano Plot -", ct),
    x = "Spatial effect (Test_2 statistic)",
    y = expression(-log[10]("FDR (BH)"))
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(linetype = "dotted", color = "grey85")
  )
ggsave("STANCE_Tcells_volcano_plot.png", p1, width = 10, height = 8, dpi = 300)
```

```{r}

```

visualization venn diagram

```{r}
library(CELINA)
library(SeuratObject)
library(sp)
load("CELINA_result.rdata", verbose = TRUE)

```

```{r}
names_in_file <- load("CELINA_result.rdata")
names_in_file
```

```{r}
class(Obj@result)
str(Obj@result, max.level = 1)
genes <- Obj@genes_list[["T-cells"]]         # grab gene names
pvals <- Obj@result[["T-cells"]]$CombinedPvals  # final combined p-values
adj_p  <- p.adjust(pvals, method = "BH")     # BH-adjusted

celina_Tcells <- data.frame(gene = genes, p_adj = adj_p)
celina_Tcells <- celina_Tcells[order(celina_Tcells$p_adj), ]  # sort smallest first

head(celina_Tcells, 20)   # show top 20
sig_celina <- subset(celina_Tcells, p_adj < 0.05)
celina_top200 <- head(sig_celina, 200)
saveRDS(celina_top200, file = "CELINA_Tcells_top200.rds")
# celina_top200 <- readRDS("CELINA_Tcells_top200.rds")

```

moran's I

```{r}

library(Seurat)
library(SeuratObject)
# library(Matrix)
# load in data
load(paste0("/Users/nikkixiaomengqi/Desktop/25_Summer_Research/breast_cancer_visium_img_updated.rdata"))
raw_counts <- GetAssayData(visium_data, assay = "RNA", slot = "counts")

# top features using RNA
DefaultAssay(visium_data) <- "RNA"

visium_data <- NormalizeData(visium_data, verbose = FALSE)
visium_data <- FindVariableFeatures(visium_data, verbose = FALSE)

visium_data <- FindSpatiallyVariableFeatures(
  visium_data,
  assay = "RNA",
  slot = "data",                         # use "data" for RNA
  features = VariableFeatures(visium_data)[1:1000],
  selection.method = "moransi",
  x.cuts = 100, y.cuts = 100
)


```

```{r}
moransi_top200 <- head(SpatiallyVariableFeatures(visium_data, method = "moransi"), 200)
moransi_top200
```

```{r}
# stance top 200
stance_top200 <- head(df, 200)
# stance_top200
```

STANCE test 1 top 200 genes

```{r}
df <- mySTANCE@Test_1  # change to your cell type if different
# head(df)
# stance_top200 <- head(df[order(df$p_value), ], 800)
stance_top200 <- df[order(df$p_value_adj, -df$statistic), ][1:200, ]
stance_top200
summary(df$p_value_adj)
```

Venn

```{r}
library(ggvenn)
stance_genes_top200 <- rownames(stance_top200)

sets <- list(
  CELINA = as.character(celina_top200$gene),
  STANCE = as.character(stance_genes_top200),
  MORANSI  = as.character(moransi_top200)
)

library(VennDiagram)
vp <- venn.diagram(
  x = sets,
  filename = NULL,                      # <- draw to screen instead of file
  fill = c("#A6CEE3", "#B2DF8A", "#FB9A99"),
  alpha = 0.6, cex = 1.2, cat.cex = 1.2,
  
  cat.pos  = c(330, 30, 180),            # CELINA left, STANCE right, MORAN bottom
  cat.dist = c(0.05, 0.05, 0.05),       # push labels slightly away from circles
  cat.just = list(c(0.5,0.5), c(0.5,0.5), c(0.5,0.5))
)
grid.newpage(); grid.draw(vp)
png("NEW_venn_top200.png", width = 1800, height = 1600, res = 300)
grid.draw(vp)
dev.off()
```

```{r}

```
